\textbf{SPECFEM3D}\\ \textbf{User Manual}

### 2010-11-11 doc/USER_MANUAL/manual_SPECFEM3D.tex
updates SPECFEM3D manual; modified output path for adjoint seismograms to have same behavior as for forward seismograms; updates some utils scripts which referred to older version

\begin{description}
\item [{\texttt{LOCAL\_PATH\_IS\_ALSO\_GLOBAL}}] Set to \texttt{.false.}
on most cluster applications. For reasons of speed, the (parallel)
mesher typically writes a (parallel) database for the solver on the
local disks of the compute nodes. Some systems have no local disks,
e.g., BlueGene or the Earth Simulator, and other systems have a fast
parallel file system, in which case this flag should be set to \texttt{.true.}.
Note that this flag is not used by the mesher or the solver; it is
only used for some of the post-processing.
\end{description}
The package can run either in single or in double precision. The default
is single precision mode because this requires exactly half as much
memory. Select your preference by selecting the appropriate setting
in the \texttt{constants.h} file:

@@

\caption{Screenshots of the CUBIT examples provided in subdirectory  \texttt{examples/}: homogeneous halfspace, layered halfspace, waterlayered halfspace and tomographic model.}

@@

\subsection{Exporting the Mesh with \texttt{cubit2specfem3d.py} }

@@

For an elastic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material. 
\item [Vp] P-Velocity of the material (given in m/s).
\item [Vs] S-Velocity of the material (given in m/s).
\item [rho] density of the material (given in kg/$m^3$).
\item [Q] quality factor to use in case of a simulation with attenuation turned on. It should be between 1 and 9000. In case no attenuation information is available, it can be set to zero. Please note, that the Vp- and Vs-velocities are given for a reference frequency. To change this reference frequency, you change the value of 
\texttt{ATTENUATION\_f0\_REFERENCE} in the main constants file \texttt{constants.h} found in subdirectory \texttt{src/shared/}.
\item [anisotropic\_flag] Flag describing the anisotropic model to use in case an anisotropic simulation should be conducted. See the file \texttt{model\_aniso.f90} in subdirectory \texttt{src/generate\_databases/} for an implementation of the anisotropic models. In case no anisotropy is available, it can be set to zero.
\end{description}
Note that this material block has to be defined using all the volumes which belong to this elastic material. For volumes belonging to another, different material, you will need to define a new material block.

@@

After the block definitions are done, you export the mesh using the script \texttt{cubit2specfem3d.py} provided in each of the example directories. If the export was successful, you should find the following files a the subdirectory \texttt{MESH/}:
\begin{description}
\item [nummaterial\_velocity\_file] Defines the material properties. For fully defined materials, the formats is:
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{rho}  \texttt{vp}  \texttt{vs}  \texttt{Q}  \texttt{anisotropy\_flag} 
\end{lyxcode}
where 
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a unique identifier,  \texttt{rho} the density in $kg \, m^{-3}$,  \texttt{vp} the P-velocity in $m \, s^{-1}$, \texttt{Vs}  the S-velocity in $m \, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag} an identifier for anisotropic models.
For a tomographic model, the material definition format is: 
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{tomography}  \texttt{name}  
\end{lyxcode}
where
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a negative, unique identifier, \texttt{tomography} keyword for tomographic material definition and \texttt{name} the name of the tomography file. The name is not used so far, rather change the filename defined in the file ~\\
\texttt{src/generate\_databases/model\_tomography.f90}.
\item [materials\_file] Contains the material associations for each element. 
\item [nodes\_coords\_file] Contains the point locations in Cartesian coordinates of the mesh element corners.
\item [mesh\_file] Contains the mesh element connectivity.
\item [free\_surface\_file] Contains the free surface connectivity.
\item [absorbing\_surface\_file\_xmax] Contains the surface connectivity of the absorbing boundary surface at the Xmax.
\item [absorbing\_surface\_file\_xmin] Contains the surface connectivity of the absorbing boundary surface at the Xmin.
\item [absorbing\_surface\_file\_ymax] Contains the surface connectivity of the absorbing boundary surface at the Ymax.
\item [absorbing\_surface\_file\_ymin] Contains the surface connectivity of the absorbing boundary surface at the Ymin.
\item [absorbing\_surface\_file\_bottom] Contains the surface connectivity of the absorbing boundary surface at the bottom.
\end{description}

@@

The partitioning is done in serial for now (in the next release we will provide a parallel version of that code),
the synopsis is:
\begin{lyxcode}
./bin/xdecompose\_mesh\_SCOTCH  nparts  input\_directory output\_directory
\end{lyxcode}

All the files generated by the Python scripts must be placed in the \texttt{input\_directory} folder before running the program.

@@

\noindent \begin{flushleft}
%
\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[scale=0.5]{figures/socal_map_mpi}
\par\end{centering}

@@

\par\end{flushleft}

@@

\item [{\texttt{LOCAL\_PATH}}] Directory in which the databases generated

@@

@@

The mesher generates the necessary databases in parallel, one set

@@

@@

@@

     \item \texttt{Q}           : 0=no attenuation / standard \texttt{Q} attenutation value

@@

In the main directory type '\texttt{make generate\_databases}'. Input for the program is provided through 
the main parameter file \texttt{Par\_file}, which resides in the subdirectory \texttt{in\_data\_files/}. Please note that  
\texttt{xgenerate\_databases} must be called directly from the \texttt{bin/} directory, as most of the binaries of the package.
Before running \texttt{xgenerate\_databases}, a number of parameters need to be set in the \texttt{Par\_file}:

\begin{description}
\item [{\texttt{SIMULATION\_TYPE}}] is set to 1 for forward simulations,
2 for adjoint simulations (see Section \ref{sec:Adjoint-simulation-finite})
and 3 for kernel simulations (see Section \ref{sec:Finite-Frequency-Kernels}).
\item [{\texttt{SAVE\_FORWARD}}] is only set to \texttt{.true.} for a forward
simulation with the last frame of the simulation saved, as part of
the finite-frequency kernel calculations (see Section \ref{sec:Finite-Frequency-Kernels}).
For a regular forward simulation, leave \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD} at their default values.
\item [{\texttt{\noun{UTM\_PROJECTION\_ZONE}}}] UTM projection zone in
which your model resides, only valid when \texttt{SUPPRESS\_UTM\_}~\\
\texttt{PROJECTION} is \texttt{.false.}.
\item [{\texttt{SUPPRESS\_UTM\_PROJECTION}}] set to be \texttt{.false.}
when your model range is specified in the geographical coordinates,
and needs to be \texttt{.true.} when your model is specified in a
cartesian coordinates. \noun{UTM projection zone in which your simulation
region resides.}
\item [{\texttt{NPROC}}] The number of MPI processors, each one is assigned one slice of the whole mesh.
\item [{\texttt{NSTEP}}] The number of time steps of the simulation. This controls the
length of the numerical simulation, i.e., twice the number of time steps
requires twice as much CPU time. This feature is not used at the time
of meshing but is required for the solver, i.e., you may change this
parameter after running  \texttt{xgenerate\_databases}.
\item [{\texttt{DT}}] The length of each time step in seconds.
If you used xmeshfem3D to generate your mesh, you should set the value suggested in \texttt{in\_out\_files/OUTPUT\_FILES/output\_mesher.txt}.

@@

\item [{\texttt{TOPOGRAPHY}}] Set to \texttt{.true.} if topography and
bathymetry should be incorporated based upon model ETOPO5 \citep{Etopo5}.
This feature adds no cost to the simulation.

@@

@@

@@

\item [{\texttt{RECORD\_LENGTH\_IN\_MINUTES}}] Choose the desired record
length of the synthetic seismograms (in minutes). This controls the
length of the numerical simulation, i.e., twice the record length
of meshing but is required for the solver, i.e., you may change this

@@

@@

\item [{\texttt{LOCAL\_PATH}}] Directory in which the databases and the synthetics

@@

@@

@@

@@

\begin{itemize}
\item the simulation type control parameters: \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD}
\item the time step parameters \texttt{NSTEP} and \texttt{DT}
\item the movie control parameters \texttt{MOVIE\_SURFACE}, \texttt{MOVIE\_VOLUME},
and \texttt{NTSTEPS\_BETWEEN\_FRAMES}
\item the ShakeMap\textregistered option \texttt{CREATE\_SHAKEMAP}
\item the output information parameters \texttt{NTSTEP\_BETWEEN\_OUTPUT\_INFO}
and \texttt{NTSTEP\_BETWEEN\_OUTPUT\_}~\\
\texttt{SEISMOS}
\item the \texttt{PRINT\_SOURCE\_TIME\_FUNCTION} flags
\end{itemize}
Any other change to the \texttt{Par\_file} implies rerunning both
the generate\_databases and the solver.

@@

\begin{lyxcode}
{\small Station~Network~Latitude~(degrees)~Longitude~(degrees)~Elevation~(m)~burial~(m)~}{\small \par}
\end{lyxcode}
The mesher filters the list of stations in file \texttt{in\_data\_files/STATIONS}
to exclude stations that are not located within the region given in
the \texttt{Par\_file} (between \texttt{LATITUDE\_MIN} and \texttt{LATITUDE\_MAX}
and between \texttt{LONGITUDE\_MIN} and \texttt{LONGITUDE\_MAX}).
The filtered file is called \texttt{in\_data\_files/STATIONS\_FILTERED}.

@@

Having set the \texttt{Par\_file} parameters, and having provided
the \texttt{CMTSOLUTION} and \texttt{STATIONS} files, you are now
ready to launch the solver! This is most easily accomplished based
upon the \texttt{go\_solver} script (See Chapter \ref{cha:Scheduler}
for information about running through a scheduler, e.g., LSF). You
may need to edit the last command at the end of the script that invokes
the \texttt{mpirun} command. The \texttt{runall} script compiles and
runs both mesher and solver in sequence. This is a safe approach that
ensures using the correct combination of mesher output and solver
input.

@@

For the same model, one can rerun the solver for different events
by simply changing the \texttt{CMTSOLUTION} file, or for different
stations by changing the \texttt{STATIONS} file. There is no need
to rerun the mesher. Of course it is best to include as many stations
as possible, since this does not add to the cost of the simulation.

@@

\begin{itemize}
\item The processors/nodes to be used for each run are assigned dynamically
by the scheduler, based on availability. Therefore, in order for the
mesher and the solver (or between successive runs of the solver) to
have access to the same database files (if they are stored on hard
drives local to the nodes on which the code is run), they must be
launched in sequence as a single job.
\item On some systems, the nodes to which running jobs are assigned are
not configured for compilation. It may therefore be necessary to pre-compile
both the mesher and the solver. 
%obsolete
%A small program provided in the distribution
%called \texttt{\small create\_header\_file.f90} can be used to directly
%create \texttt{\small in\_out\_files/OUTPUT\_FILES/values\_from\_mesher.h} using
%the information in the \texttt{\small in\_data\_files/Par\_file} without having
%to run the mesher (type \texttt{\small `make}{\small{} }\texttt{\small create\_header\_}~\\
%\texttt{\small file}' to compile it and `\texttt{\small xcreate\_header\_file}'
%to run it; refer to the sample scripts below). The solver can now
%be compiled as explained above.
\item One feature of schedulers/queuing systems is that they allow submission
of multiple jobs in a ``launch and forget'' mode. In order to take
advantage of this property, care needs to be taken that output and
intermediate files from separate jobs do not overwrite each other,
or otherwise interfere with other running jobs.
\end{itemize}
We describe here in some detail a job submission procedure for the
Caltech 1024-node cluster, CITerra, under the LSF scheduling system.
We consider the submission of a regular forward simulation. The two
main scripts are \texttt{\small run\_lsf.bash}, which compiles the
Fortran code and submits the job to the scheduler, and \texttt{\small go\_mesher\_solver\_lsf}~\\
\texttt{\small .bash}, which contains the instructions that make up
the job itself. These scripts can be found in \texttt{\small UTILS/}
directory and can straightforwardly be modified and adapted to meet
more specific running needs.

\section{\texttt{run\_lsf.bash}}

This script first sets the job queue to be `normal'. It then compiles
the mesher and solver together, figures out the number of processors
required for this simulation from the \texttt{in\_data\_files/Par\_file}, and
submits the LSF job.

@@

@@

echo~\char`\"{}Finished~compilation~\$d\char`\"{}~\\
~\\
\#~compute~total~number~of~nodes~needed~

NPROC=`grep~NPROC~in\_data\_files/Par\_file~|~cut~-c~34-~'~

~\\
\#~total~number~of~nodes~is~the~product~of~the~values~read~

numnodes=\$((~\$NPROC))~\\
~\\
echo~\char`\"{}Submitting~job\char`\"{}~

bsub~\$queue~-n~\$numnodes~-W~60~-K~<go\_mesher\_solver\_lsf.bash~
\end{lyxcode}

\section{\texttt{go\_mesher\_solver\_lsf.bash}}

This script describes the job itself, including setup steps that can
only be done once the scheduler has assigned a job-ID and a set of
compute nodes to the job, the \texttt{run\_lsf.bash} commands used
to run the mesher and the solver, and calls to scripts that collect
the output seismograms from the compute nodes and perform clean-up
operations.

\begin{enumerate}
\item First the script directs the scheduler to save its own output and
output from \texttt{stdout} into ~\\
\texttt{\small in\_out\_files/OUTPUT\_FILES/\%J.o},
where \texttt{\%J} is short-hand for the job-ID; it also tells the
scheduler what version of \texttt{mpich} to use (\texttt{mpich\_gm})
and how to name this job (\texttt{go\_mesher\_solver\_lsf}).
\item The script then creates a list of the nodes allocated to this job
by echoing the value of a dynamically set environment variable \texttt{LSB\_MCPU\_HOSTS}
and parsing the output into a one-column list using the Perl script
\texttt{UTILS/remap\_lsf\_machines.pl}. It then creates a set of scratch
directories on these nodes (\texttt{\small /scratch/}~\\
\texttt{\small \$USER/DATABASES\_MPI}) to be used as the \texttt{LOCAL\_PATH}
for temporary storage of the database files. The scratch directories
are created using \texttt{shmux}, a shell multiplexor that can execute
the same commands on many hosts in parallel. \texttt{shmux} is available
from Shmux \url{web.taranis.org/shmux/}. Make sure that the \texttt{LOCAL\_PATH}
parameter in \texttt{in\_data\_files/Par\_file} is also set properly.
\item The next portion of the script launches the mesher and then the solver
using \texttt{run\_lsf.bash}.
\item The final portion of the script collects the seismograms and performs
clean up on the nodes, using the Perl scripts \texttt{collect\_seismo\_lsf\_multi.pl}
and \texttt{cleanmulti.pl}.
\end{enumerate}
\begin{lyxcode}
\#!/bin/bash~-v~

@@

mkdir~-p~in\_out\_files/OUTPUT\_FILE

>>>>

\begin{description}
\item [{\texttt{LOCAL\_PATH\_IS\_ALSO\_GLOBAL}}] Set to \texttt{.false.}
on most cluster applications. For reasons of speed, the (parallel)
distributed database generator typically writes a (parallel) database for the solver on the
local disks of the compute nodes. Some systems have no local disks,
e.g., BlueGene or the Earth Simulator, and other systems have a fast
parallel file system, in which case this flag should be set to \texttt{.true.}.
Note that this flag is not used by the database generator or the solver; it is
only used for some of the post-processing.
\end{description}
The package can run either in single or in double precision. The default
is single precision mode because this requires exactly half as much
memory. Select your preference by selecting the appropriate setting
in the \texttt{constants.h} file:

@@

\caption{Screenshots of the CUBIT examples provided in subdirectory  \texttt{examples/}: homogeneous halfspace (top-left), layered halfspace (top-right), waterlayered halfspace (bottom-left) and tomographic model (bottom-right).}

@@

\subsection{Exporting the Mesh with \texttt{cubit2specfem3d.py} }
\label{subsec:Exporting-the-Mesh}

@@

For an elastic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material. 
\item [Vp] P-Velocity of the material (given in m/s).
\item [Vs] S-Velocity of the material (given in m/s).
\item [rho] density of the material (given in kg/$m^3$).
\item [Q] quality factor to use in case of a simulation with attenuation turned on. It should be between 1 and 9000. In case no attenuation information is available, it can be set to zero. Please note that your Vp- and Vs-velocities are given for a reference frequency. To change this reference frequency, you change the value of 
\texttt{ATTENUATION\_f0\_REFERENCE} in the main constants file \texttt{constants.h} found in subdirectory \texttt{src/shared/}.
\item [anisotropic\_flag] Flag describing the anisotropic model to use in case an anisotropic simulation should be conducted. See the file \texttt{model\_aniso.f90} in subdirectory \texttt{src/generate\_databases/} for an implementation of the anisotropic models. In case no anisotropy is available, it can be set to zero.
\end{description}
Note that this material block has to be defined using all the volumes which belong to this elastic material. For volumes belonging to another, different material, you will need to define a new material block.

@@

After the block definitions are done, you export the mesh using the script \texttt{cubit2specfem3d.py} provided in each of the example directories. If the export was successful, you should find the following files in a subdirectory \texttt{MESH/}:

\begin{description}
\item [nummaterial\_velocity\_file] Defines the material properties. For fully defined materials, the formats is:
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{rho}  \texttt{vp}  \texttt{vs}  \texttt{Q}  \texttt{anisotropy\_flag} 
\end{lyxcode}
where 
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a unique identifier,  \texttt{rho} the density in $kg \, m^{-3}$,  \texttt{vp} the P-velocity in $m \, s^{-1}$, \texttt{vs}  the S-velocity in $m \, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag} an identifier for anisotropic models.
For a tomographic model, the material definition format is: 
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{tomography}  \texttt{name}  
\end{lyxcode}
where
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a negative, unique identifier, \texttt{tomography} keyword for tomographic material definition and \texttt{name} the name of the tomography file. The name is not used so far, rather change the filename defined in the file \texttt{model\_tomography.f90} located in the
\texttt{src/generate\_databases/} directory.
\item [materials\_file] Contains the material associations for each element. 
\item [nodes\_coords\_file] Contains the point locations in Cartesian coordinates of the mesh element corners.
\item [mesh\_file] Contains the mesh element connectivity.
\item [free\_surface\_file] Contains the free surface connectivity.
\item [absorbing\_surface\_file\_xmax] Contains the surface connectivity of the absorbing boundary surface at the Xmax.
\item [absorbing\_surface\_file\_xmin] Contains the surface connectivity of the absorbing boundary surface at the Xmin.
\item [absorbing\_surface\_file\_ymax] Contains the surface connectivity of the absorbing boundary surface at the Ymax.
\item [absorbing\_surface\_file\_ymin] Contains the surface connectivity of the absorbing boundary surface at the Ymin.
\item [absorbing\_surface\_file\_bottom] Contains the surface connectivity of the absorbing boundary surface at the bottom.
\end{description}

@@

The partitioning is done in serial for now (in the next release we will provide a parallel version of that code),
the synopsis is:
\begin{lyxcode}
./bin/xdecompose\_mesh\_SCOTCH  nparts  input\_directory output\_directory
\end{lyxcode}
where
\begin{itemize}
     \item \texttt{nparts} is the number of partitions, i.e. number of cores for the parallel simulations,
     \item \texttt{input\_directory} is the directory which holds all the files generated by the Python script \texttt{cubit2specfem3d.py} explained in the previous Section~\ref{subsec:Exporting-the-Mesh}, e.g. \texttt{MESH/}, and
     \item \texttt{output\_directory} is the directory for the output of this partitioner which stores ACII-format files named like \texttt{proc******\_Database} for each partition. These files will be needed for creating the distributed databases, and have to reside in the directory  \texttt{LOCAL\_PATH} specified in the main \texttt{Par\_file}, e.g. in directory \texttt{in\_out\_files/DATABASES\_MPI}. Please see Chapter~\ref{cha:Creating-Distributed-Databases} for further details.
\end{itemize}
Note that all the files generated by the Python script  \texttt{cubit2specfem3d.py} must be placed in the \texttt{input\_directory} folder before running the program.

@@

%\noindent \begin{flushleft}
%
\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[scale=0.5]{figures/socal_map_mpi}
\par\end{centering}

@@

%\par\end{flushleft}

@@

\item [{\texttt{LOCAL\_PATH}}] Directory in which the partitions generated

@@

local disk of the compute nodes, although on some machines these partitions

@@

The mesher generates the necessary partitions in parallel, one set

@@

the contents of the \texttt{LOCAL\_PATH} directory to see the

@@

These files will be needed for creating the distributed databases, and have to reside in the directory  \texttt{LOCAL\_PATH} specified in the main \texttt{Par\_file}, e.g. in directory \texttt{in\_out\_files/DATABASES\_MPI}. Please see Chapter~\ref{cha:Creating-Distributed-Databases} for further details.

@@

     \item \texttt{Q}           : 0=no attenuation / standard \texttt{Q} attenuation value

@@

In the main directory type '\texttt{make generate\_databases}'. Input for the program is provided through 
the main parameter file \texttt{Par\_file}, which resides in the subdirectory \texttt{in\_data\_files/}. Please note that  
\texttt{xgenerate\_databases} must be called directly from the \texttt{bin/} directory, as most of the binaries of the package.

\section{\label{cha:Main-Parameter}Main parameter file \texttt{Par\_file}}

Before running \texttt{xgenerate\_databases}, a number of parameters need to be set in the main parameter \texttt{Par\_file}
located in the subdirectory \texttt{in\_data\_files/}:

\begin{description}
\item [{\texttt{SIMULATION\_TYPE}}] is set to 1 for forward simulations,
2 for adjoint simulations (see Section \ref{sec:Adjoint-simulation-finite})
and 3 for kernel simulations (see Section \ref{sec:Finite-Frequency-Kernels}).
\item [{\texttt{SAVE\_FORWARD}}] is only set to \texttt{.true.} for a forward
simulation with the last frame of the simulation saved, as part of
the finite-frequency kernel calculations (see Section \ref{sec:Finite-Frequency-Kernels}).
For a regular forward simulation, leave \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD} at their default values.
\item [{\texttt{\noun{UTM\_PROJECTION\_ZONE}}}] UTM projection zone in
which your model resides, only valid when \texttt{SUPPRESS\_UTM\_}~\\
\texttt{PROJECTION} is \texttt{.false.}.
\item [{\texttt{SUPPRESS\_UTM\_PROJECTION}}] set to be \texttt{.false.}
when your model range is specified in the geographical coordinates,
and needs to be \texttt{.true.} when your model is specified in a
cartesian coordinates. \noun{UTM projection zone in which your simulation
region resides.}
\item [{\texttt{NPROC}}] The number of MPI processors, each one is assigned one slice of the whole mesh.
\item [{\texttt{NSTEP}}] The number of time steps of the simulation. This controls the
length of the numerical simulation, i.e., twice the number of time steps
of generating the distributed databases but is required for the solver, i.e., you may change this
\item [{\texttt{DT}}] The length of each time step in seconds. 
This feature is not used at the time of generating the distributed databases but is required for the solver.
Please see also Section~\ref{cha:Choosing-the-Time-Step} for further details.

@@

%obsolete...
%\item [{\texttt{TOPOGRAPHY}}] Set to \texttt{.true.} if topography and
%bathymetry should be incorporated based upon model ETOPO5 \citep{Etopo5}.
%This feature adds no cost to the simulation.

@@

Please note that the Vp- and Vs-velocities of your model are given for a reference frequency. 
To change this reference frequency, you change the value of 
\texttt{ATTENUATION\_f0\_REFERENCE} in the main constants file \texttt{constants.h} found in subdirectory \texttt{src/shared/}.

@@

\item [{\texttt{ANISOTROPY}}] Set to \texttt{.true.} if you
want to use an anisotropy model. Please see the file \texttt{model\_aniso.f90} in subdirectory \texttt{src/generate\_databases/} for the current implementation of anisotropic models.

@@

%obsolete...
%\item [{\texttt{RECORD\_LENGTH\_IN\_MINUTES}}] Choose the desired record
%length of the synthetic seismograms (in minutes). This controls the
%length of the numerical simulation, i.e., twice the record length
%requires twice as much CPU time. This feature is not used at the time
%of meshing but is required for the solver, i.e., you may change this
%parameter after running \texttt{xgenerate\_databases}.

@@

\item [{\texttt{HDUR\_MOVIE}}] determines the half duration of the source time function for the movie simulations. 
When this parameter is set to be 0, a default half duration that corresponds to the accuracy of the simulation is provided. 
Otherwise, it adds this half duration to the half duration specified in the source file \texttt{CMTSOLUTION},
thus simulates longer periods to make the movie images look smoother.

@@

\item [{\texttt{LOCAL\_PATH}}] Directory in which the distributed databases

@@

for each of the \texttt{NPROC} slices that constitutes
the mesh (see Figure~\ref{fig:mount.partitions} and Figure~\ref{fig:For-parallel-computing}). After
the executable finishes, you can log in to one of the compute nodes and view

@@

Please note that the \texttt{LOCAL\_PATH} directory should already contain the output files of 
the partitioner, i.e. from \texttt{xdecompose\_mesh\_SCOTCH} or \texttt{xmeshfem3D}.

@@

\section{\label{cha:Choosing-the-Time-Step}Choosing the time step \texttt{DT}}

The parameter \texttt{DT} sets the length of each time step in seconds. The value of this parameter is crucial for the stability of 
the spectral-element simulation. Your time step \texttt{DT} will depend on the minimum ratio between the distance $h$ of neighboring mesh points and the wave speeds $v$ defined in your model. The condition for the time step $\Delta t$ is:

\begin{lyxcode}
$\Delta t < C ~\mathrm{min}_\Omega (~ h / v ~)$
\end{lyxcode}
where $C$ is the so-called Courant number and $\Omega$ denotes the model volume. 
The distance $h$ depends on the mesh element size and the number of
GLL points \texttt{NGLL} specified in the main constants file \texttt{constants.h} located in the \texttt{src/shared/} subdirectory.
The wave speed $v$ is determined based on your model's P- (or S-) velocity values.

The database generator \texttt{xgenerate\_databases}, as well as the internal mesher \texttt{xmeshfem3D}, are trying
to evaluate the value of $\Delta t$ for empirically chosen Courant numbers $C \sim 0.3$.
If you used the mesher \texttt{xmeshfem3D} to generate your mesh, you should set the value suggested in \texttt{in\_out\_files/OUTPUT\_FILES/output\_mesher.txt} file, which is created after the mesher completed. In case you used CUBIT to create the mesh, you might use an arbitrary value when running \texttt{xgenerate\_databases} and then use the value suggested in the ~\\
\texttt{in\_out\_files/OUTPUT\_FILES/output\_mesher.txt} file after the database generation completed. 
Note that the implemented Newmark time scheme uses this time step globally, thus your simulations become more expensive for very small mesh elements in high wave-speed regions. Please be aware of this restriction when constructing your mesh in Chapter~\ref{cha:Mesh-Generation}.

@@

\begin{itemize}
\item the simulation type control parameters: \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD}
\item the time step parameters \texttt{NSTEP} and \texttt{DT}
\item the attenuation control parameters \texttt{ATTENUATION} and \texttt{USE\_OLSEN\_ATTENUATION}
\item the absorbing boundary control parameter \texttt{ABSORBING\_CONDITIONS} 
\item the movie control parameters \texttt{MOVIE\_SURFACE}, \texttt{MOVIE\_VOLUME},
and \texttt{NTSTEPS\_BETWEEN\_FRAMES}
\item the ShakeMap\textregistered option \texttt{CREATE\_SHAKEMAP}
\item the output information parameters \texttt{NTSTEP\_BETWEEN\_OUTPUT\_INFO}
and \texttt{NTSTEP\_BETWEEN\_OUTPUT\_}~\\
\texttt{SEISMOS}
\item the \texttt{PRINT\_SOURCE\_TIME\_FUNCTION} flags
\end{itemize}
Any other change to the \texttt{Par\_file} implies rerunning both
the database generator \texttt{xgenerate\_databases} and the solver \texttt{xspecfem3D}.

@@

\begin{lyxcode}
{\small Station~Network~Latitude~(degrees)~Longitude~(degrees)~Elevation~(m)~burial~(m)~}{\small \par}
\end{lyxcode}
The solver \texttt{xspecfem3D} filters the list of stations in file \texttt{in\_data\_files/STATIONS}
to exclude stations that are not located within the region given in
the \texttt{Par\_file} (between \texttt{LATITUDE\_MIN} and \texttt{LATITUDE\_MAX}
and between \texttt{LONGITUDE\_MIN} and \texttt{LONGITUDE\_MAX}).
The filtered file is called \texttt{in\_data\_files/STATIONS\_FILTERED}.

@@

Having set the \texttt{Par\_file} parameters, and having provided
the \texttt{CMTSOLUTION} and \texttt{STATIONS} files, you are now
ready to launch the solver! This is most easily accomplished based
upon the \texttt{go\_solver} script (See Chapter \ref{cha:Scheduler}
for information about running through a scheduler, e.g., LSF). You
may need to edit the last command at the end of the script that invokes
the \texttt{mpirun} command. The \texttt{runall} script compiles and
runs both \texttt{xgenerate\_databases} and \texttt{xspecfem3D} in sequence. This is a safe approach that
ensures using the correct combination of distributed database output and solver
input.

@@

For the same model, one can rerun the solver for different events
by simply changing the \texttt{CMTSOLUTION} file, or for different
stations by changing the \texttt{STATIONS} file. There is no need
to rerun the \texttt{xgenerate\_databases} executable. Of course it is best to include as many stations
as possible, since this does not add to the cost of the simulation.

@@

\begin{itemize}
\item The processors/nodes to be used for each run are assigned dynamically
by the scheduler, based on availability. Therefore, in order for the
\texttt{xgenerate\_databases} and the \texttt{xspecfem3D} executables (or between successive runs of the solver) to
have access to the same database files (if they are stored on hard
drives local to the nodes on which the code is run), they must be
launched in sequence as a single job.
\item On some systems, the nodes to which running jobs are assigned are
not configured for compilation. It may therefore be necessary to pre-compile
both the \texttt{xgenerate\_databases} and the \texttt{xspecfem3D} executables. 
%obsolete
%A small program provided in the distribution
%called \texttt{\small create\_header\_file.f90} can be used to directly
%create \texttt{\small in\_out\_files/OUTPUT\_FILES/values\_from\_mesher.h} using
%the information in the \texttt{\small in\_data\_files/Par\_file} without having
%to run the mesher (type \texttt{\small `make}{\small{} }\texttt{\small create\_header\_}~\\
%\texttt{\small file}' to compile it and `\texttt{\small xcreate\_header\_file}'
%to run it; refer to the sample scripts below). The solver can now
%be compiled as explained above.
\item One feature of schedulers/queuing systems is that they allow submission
of multiple jobs in a ``launch and forget'' mode. In order to take
advantage of this property, care needs to be taken that output and
intermediate files from separate jobs do not overwrite each other,
or otherwise interfere with other running jobs.
\end{itemize}
Examples of job scripts can be found in \texttt{\small UTILS/Cluster/}
directory and can straightforwardly be modified and adapted to meet
more specific running needs.

We describe here in some detail a job submission procedure for the
Caltech 1024-node cluster, CITerra, under the LSF scheduling system.
We consider the submission of a regular forward simulation using the internal mesher
to create mesh partitions. The two
main scripts are \texttt{\small run\_lsf.bash}, which compiles the
Fortran code and submits the job to the scheduler, and \texttt{\small go\_mesher\_solver\_lsf\_basin.forward}, 
which contains the instructions that make up
the job itself. These scripts can be found in \texttt{\small UTILS/Cluster/lsf/} directory

\section{Job submission \texttt{run\_lsf.bash}}

This script first sets the job queue to be `normal'. It then compiles
the mesher, database generator and solver together, figures out the number of processors
required for this simulation from the \texttt{in\_data\_files/Par\_file}, and
submits the LSF job.

@@

make~meshfem3D~

@@

echo~\char`\"{}Finished~compilation~\$d\char`\"{}~\\
~\\
\#~get~total~number~of~nodes~needed~for~solver~

NPROC=`grep~NPROC~in\_data\_files/Par\_file~|~cut~-c~34-~'~\\
~\\

\#~compute~total~number~of~nodes~needed~for~mesher~

NPROC\_XI=`grep~NPROC\_XI~in\_data\_files/meshfem3D\_files/Mesh\_Par\_file~|~cut~-c~34-~'~
~\\

NPROC\_ETA=`grep~NPROC\_ETA~in\_data\_files/meshfem3D\_files/Mesh\_Par\_file~|~cut~-c~34-~'~
~\\
\#~total~number~of~nodes~is~the~product~of~the~values~read~

numnodes=\$((~\$NPROC\_XI~*~\$NPROC\_ETA~))~\\
~\\

\#~checks~total~number~of~nodes~

if~{[}~\$numnodes~-neq~\$NPROC~];~then

~~~~~~~~echo~\char`\"{}error number of procs mismatch\char`\"{}

~~~~~~~~exit~

fi~\\

~\\
echo~\char`\"{}Submitting~job\char`\"{}~

bsub~\$queue~-n~\$numnodes~-W~60~-K~<go\_mesher\_solver\_lsf.forward~
\end{lyxcode}

\section{Job script \texttt{go\_mesher\_solver\_lsf.forward}}

This script describes the job itself, including setup steps that can
only be done once the scheduler has assigned a job-ID and a set of
compute nodes to the job, the \texttt{run\_lsf.bash} commands used
to run the mesher, database generator and the solver, and calls to scripts that collect
the output seismograms from the compute nodes and perform clean-up
operations.

\begin{enumerate}
\item First the script directs the scheduler to save its own output and
output from \texttt{stdout} into ~\\
\texttt{\small in\_out\_files/OUTPUT\_FILES/\%J.o},
where \texttt{\%J} is short-hand for the job-ID; it also tells the
scheduler what version of \texttt{mpich} to use (\texttt{mpich\_gm})
and how to name this job (\texttt{go\_mesher\_solver\_lsf}).
\item The script then creates a list of the nodes allocated to this job
by echoing the value of a dynamically set environment variable \texttt{LSB\_MCPU\_HOSTS}
and parsing the output into a one-column list using the Perl script
\texttt{UTILS/remap\_lsf\_machines.pl}. It then creates a set of scratch
directories on these nodes (\texttt{\small /scratch/}~\\
\texttt{\small \$USER/DATABASES\_MPI}) to be used as the \texttt{LOCAL\_PATH}
for temporary storage of the database files. The scratch directories
are created using \texttt{shmux}, a shell multiplexor that can execute
the same commands on many hosts in parallel. \texttt{shmux} is available
from Shmux \url{web.taranis.org/shmux/}. Make sure that the \texttt{LOCAL\_PATH}
parameter in \texttt{in\_data\_files/Par\_file} is also set properly.
\item The next portion of the script launches the mesher, database generator and then the solver
using \texttt{run\_lsf.bash}.
\item The final portion of the script collects the seismograms and performs
clean up on the nodes, using the Perl scripts \texttt{collect\_seismo\_lsf\_multi.pl}
and \texttt{cleanmulti.pl}.
\end{enumerate}
\begin{lyxcode}
\#!/bin/bash~-v~

@@

mkdir~-p~in\_out\_files/OUTPUT\_FILES~
